---
title: "Exploratory Regression"
author: "Liz Cutting"
date: "27 October 2025"
output:
  pdf_document: default
---

```{r setup, include=FALSE, message=FALSE,warning=FALSE}
library(knitr)
library(tinytex)
library(tidyverse)
library(dplyr)

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(out.width = "60%",fig.align='center')
```

```{r}
data <- read_csv("fire_data_2023.csv")

glimpse(data)

data <- na.omit(data)
data$SIZE_HA <- as.numeric(data$SIZE_HA)
h <- hist(data$SIZE_HA, plot = FALSE)   # compute histogram, but don’t plot

plot(h$mids, log(h$counts),
     type = "h", lwd = 10,
     main = "Histogram of (log(Fire size))",
     xlab = "Size (HA)",
     ylab = "log(Count)",
     col = "skyblue")


```


```{r}
model <- lm(SIZE_HA~SPECIFIC_HUMIDITY+TEMP+PRECIP_ICE+PRECIP_WATER+PRECIP_ICE+WIND+TSURF+GWETTOP+LHLAND + SHLAND+LAI+GRN+SWLAND+EVPTRNS+RZMC, data = data)
summary(model)
```
```{r}
plot(model)
```

This data assumes we have a normal distribution. But, we don't! See Q-Q plot. So, try finding outliers. When that doesn't hlep, try a log-normal by transforming the response. 


```{r}
library(ggplot2)
library(dplyr)

filtered_data <- data %>%
  filter(SIZE_HA <= 50)

filtered_data <- filtered_data %>%
  filter(SHLAND > 0)

ggplot(filtered_data, aes(x = WIND, y = SIZE_HA)) +
  geom_point()+
  geom_smooth(formula=y~x,method="lm",se=FALSE,color="blue")+
  labs(title="Association between ",
       x="Width (cm)",
       y="Weight (g)",
       caption="Source: Brofeldt (1917)") +
  theme_bw()
```


```{r}
cleaned_data <- na.omit(data)
data_cleaned <- cleaned_data %>% filter(SIZE_HA > 0)
data2 <- data_cleaned %>% 
        mutate(log_SIZE_HA = log(SIZE_HA))

glimpse(data2)

model <- lm(log_SIZE_HA~SPECIFIC_HUMIDITY+TEMP+PRECIP_ICE+PRECIP_WATER+PRECIP_ICE+WIND+TSURF+GWETTOP+LHLAND + SHLAND+LAI+GRN+SWLAND+EVPTRNS+RZMC, data = data2)
summary(model)
```

The log of the RSE above was 10.07, so this is quite a large improvement. Check normamlity conditions:

```{r}
plot(model)

```

```{r}
library(car)
vif(model)


coef_df <- tibble(
  variable = c(
    "SPECIFIC_HUMIDITY", "TEMP", "PRECIP_ICE", "PRECIP_WATER", "WIND",
    "TSURF", "GWETTOP", "LHLAND", "SHLAND", "LAI", "GRN", "SWLAND",
    "EVPTRNS", "RZMC"
  ),
  value = c(
    9.805781, 12.134880, 1.653941, 1.696050, 1.686045,
    5.933761, 33.974923, 29.080029, 11.374341, 2.897738,
    1.517025, 25.168641, 9.287284, 23.210578
  )
)


coef_df2 <- tibble(
  variable = c(
    "SPECIFIC_HUMIDITY", "TEMP", "PRECIP_ICE", "PRECIP_WATER", "WIND",
    "TSURF", "GWETTOP", "LHLAND"
  ),
  value = c(
    0.00531, 0.06649, 0.6697, 0.3788, 0.00395, 0.715, 0.0066, 0.0799
  )
)
ggplot(coef_df2, aes(x = variable, y = value)) +
  geom_col(fill = "steelblue") +
  geom_hline(yintercept = 0.1, color="red", width=4)+
  geom_hline(yintercept = 0.05, color="darkred", width=4)+
  labs(
    title = "Collinearity Analysis",
    y = "p-value",
    x = "Variable"
  ) +
  theme_bw()
ggplot(coef_df2, aes(x = variable, y = value)) +
  geom_col(fill = "steelblue") +
  geom_hline(
    data = data.frame(threshold = c(0.10, 0.05),
                      label = c("0.10 Level", "0.05 Level")),
    aes(yintercept = threshold, color = label, linetype = label),
    linewidth = 1
  ) +
  scale_color_manual(values = c("0.10 Level" = "red", 
                                "0.05 Level" = "darkred")) +
  scale_linetype_manual(values = c("0.10 Level" = "dashed",
                                   "0.05 Level" = "dotted")) +labs(
    title = "Significant Variables in MLR",
    y = "p-value",
    x = "Variable",
    color = "Threshold",
    linetype = "Threshold"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )


ggplot(coef_df, aes(x = reorder(variable, value), y = value)) +
  geom_col(fill = "steelblue") +
  geom_hline(yintercept = 12, color="red", width=4)+
  coord_flip() +
  labs(
    title = "Collinearity Analysis",
    y = "VIF Score",
    x = "Variable"
  ) +
  theme_bw()

```
Wow, much better! This should be good enough. There is a right skew, but it's not a big deal. We see that when studying Cook's distance on the outliers, we don't have many that are considered too extreme statistically. 


Let's test out full power-law method. 

```{r}
data_pl <- data_cleaned %>%
    mutate(
        log_SIZE = log(SIZE_HA),
        across(c(SPECIFIC_HUMIDITY, TEMP, PRECIP_ICE, PRECIP_WATER, WIND,
                 TSURF, GWETTOP, LHLAND, SHLAND, LAI, GRN, SWLAND, 
                 EVPTRNS, RZMC), ~ log(.x + 1), .names = "log_{col}")
    )

model_powerlaw <- lm(log_SIZE ~ ., data = select(data_pl, starts_with("log_")))
summary(model_powerlaw)
plot(model_powerlaw)
```


This is basically the same result! Since we don't see better Adjusted R^2, we wouldn't change the model. However, we do note that the outliers might be managed even better. To confirm which model is better, we will make a decision based on 10-fold CV error: 

```{r}
library(caret)

set.seed(303)

cv_control <- trainControl(method = "cv", number = 10)

cv_lognormal <- train(
  log_SIZE_HA ~ SPECIFIC_HUMIDITY + TEMP + PRECIP_ICE + PRECIP_WATER + WIND +
    TSURF + GWETTOP + LHLAND + SHLAND + LAI + GRN + SWLAND +
    EVPTRNS + RZMC,
  data = data2,
  method = "lm",
  trControl = cv_control,
  metric = "RMSE"
)
# convert to fire size units (HA)
exp(cv_lognormal$results$RMSE)


```

```{r}
#use this to make the data more similar when doing CV for accuracy
safe_log <- function(x) {
  min_x <- min(x, na.rm = TRUE)
  shift <- ifelse(min_x <= 0, abs(min_x) + 1e-6, 0)
  log(x + shift)
}

data_pl <- data_cleaned %>%
  mutate(
    log_SIZE = log(SIZE_HA),
    across(
      c(SPECIFIC_HUMIDITY, TEMP, PRECIP_ICE, PRECIP_WATER, WIND,
        TSURF, GWETTOP, LHLAND, SHLAND, LAI, GRN, SWLAND,
        EVPTRNS, RZMC),
      ~ safe_log(.x),
      .names = "log_{col}"
    )
  )

model_powerlaw <- lm(
  log_SIZE ~ .,
  data = select(data_pl, starts_with("log_"))
)

summary(model_powerlaw)

set.seed(303)

cv_powerlaw <- train(
  log_SIZE ~ .,
  data = select(data_pl, starts_with("log_")),
  method = "lm",
  trControl = cv_control,
  metric = "RMSE"
)

exp(cv_powerlaw$results$RMSE)


```
It looks like a log-normal distribution is  best! 


